---
title: "Basic Budget"
author: "Mike Dietze"
date: "December 9, 2014"
output: html_document
---

## TASKS

1. Load up data
2. Estimate any uncertainties required as inputs, but which were not provided in the data
3. Format data for JAGS
4. Specify / Estimate Priors
5. Specify / Estimate Initial Conditions for MCMC
6. Basic Code
7. Post-Processing & Convergence Checking
8. Visualization

## 0. Settings, etc

```{r}
setwd("~/Dropbox/HF C Synthesis")
library(rjags)

ciEnvelope <- function(x,ylo,yhi,...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
                                      ylo[1])), border = NA,...) 
}
```


## 1. Load up data

```{r}
tree = read.csv("Data/live trees/hf-c-trees-live.csv",header=TRUE)
tree = tree[,-grep('^X',colnames(tree))]  ## drop columns with names starting with X
```


## 2. Estimate any uncertainties required as inputs, but which were not provided in the data

### 2a. Load Allometric posteriors

## 3. Format data for JAGS

```{r}

year = tree$year
years = min(year):max(year)
nyears = length(years)
year.index = year+1-years[1]

plot = tree$plot
plots = unique(plot)
nplots = length(plots)
plot.index = as.integer(plot)

data = list(dbh=tree$dbh,n=nrow(tree),ny=nyears,yi=year.index)


```


## 4. Specify / Estimate Priors


## 5. Specify / Estimate Initial Conditions for MCMC
```{r}
x = tree$dbh
mu.ic = mean(x,na.rm=TRUE)
prec.ic = 1/var(x,na.rm=TRUE)
alpha.ic = mu.ic - tapply(x,year.index,mean,na.rm=TRUE)
tau_year.ic = 1/var(alpha.ic,na.rm=TRUE)
alpha.ic[as.numeric(names(alpha.ic))] <- alpha.ic
alpha.ic[is.na(alpha.ic)] <- rnorm(sum(is.na(alpha.ic)),0,sqrt(1/tau_year.ic))


init =  list(mu=mu.ic,tau=prec.ic,alpha.y=alpha.ic,tau_year=tau_year.ic)
```

## 6. Basic Code

```{r}
BasicBudget = "
model{
  mu ~ dnorm(20,0.001)
  tau ~ dgamma(0.1,0.1)

  tau_year ~ dgamma(0.1,0.1)
  for(y in 1:ny){
    alpha.y[y] ~ dnorm(0,tau_year)
  }

  for(i in 1:n){
    theta[i] <- mu + alpha.y[yi[i]]
    dbh[i] ~ dnorm(theta[i],tau)
  }

}
"

## compile JAGS model
j.model   <- jags.model (file = textConnection(BasicBudget),
                           data = data,
                           inits = init,
                           n.chains = 3)
## burn-in
bmcmc   <- coda.samples (model = j.model,
                              variable.names = c("mu","tau","tau_year","alpha.y"),
                              n.iter = 2000)

```


## 7. Post-Processing & Convergence Checking

```{r}
plot(bmcmc)  		## mcmc history and density plot
autocorr.plot(bmcmc)		## autocorrelation
cumuplot(bmcmc)		## quantile plot
gelman.plot(bmcmc)		## GRB statistic
summary(bmcmc)		## summary table
mu = as.data.frame(as.matrix(bmcmc))$mu
quantile(mu,c(0.025,0.5,0.975))
```

## 8. Visualization

```{r}
b.df = as.data.frame(as.matrix(bmcmc))
ci = matrix(NA,nyears,3)
for(y in 1:nyears){
  ci[y,] <- quantile(b.df$mu+b.df[,paste0("alpha.y[",y,"]")],c(0.025,0.5,0.975))
}
plot(years,ci[,2],type='n',ylim=range(ci),ylab="Mean DBH")
ciEnvelope(years,ci[,1],ci[,3],col="lightBlue")
lines(years,ci[,2],lwd=3)
```

