---
title: "Basic Budget"
author: "Mike Dietze"
date: "December 9, 2014"
output: html_document
---

## TASKS

1. Load up data
1.1 Match treering and adult data
2. Estimate any uncertainties required as inputs, but which were not provided in the data
3. Format data for JAGS
4. Specify / Estimate Priors
5. Specify / Estimate Initial Conditions for MCMC
6. Basic Codes
6.1 Run tree ring / plot fusion at a plot scale
7. Post-Processing & Convergence Checking
8. Visualization
9. Run plot -> landscape fusion

## 0. Settings, etc

```{r}
setwd("~/Dropbox/HF C Synthesis")
library(rjags)
library(dplR)
library(PEcAn.data.land)
# library(sp)
# library(foreign)
# library(lattice)
# library(ggplot2)
# library(sciplot)
# library(data.table)

ciEnvelope <- function(x,ylo,yhi,...){
  polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
                                      ylo[1])), border = NA,...) 
}
```


## 1. Load up data

```{r}
## load plot tree data
raw.tree = read.csv("Data/live trees/hf-c-trees-live_26Jan2015.csv",header=TRUE)
#tree = raw.tree[,-grep('^X',colnames(tree))]  ## drop columns with names starting with X
tree = raw.tree[-which(is.na(raw.tree$dbh)),] ## drop rows with missing DBH
tree = raw.tree[-which(raw.tree$obs.exp == "exp"),] ## drop trees from experimental manipulations

## Testing subset
tree = tree[which(tree$site %in% c("ems","hemlock","hemlock woodlot","red pine plantation","mixed hardwood","pisgah","lyford","simes","TS")),]

## load rings
folder <- "Data/Tree_Rings/Lyford_Data_13m/RW/Combined/"
raw.rings <- Read_Tuscon(folder)
```

## 1.1 Match treerings and adult plot data
```{r}
# build tree ring codes
if(is.list(raw.rings)){
  rings <- combine.rwl(raw.rings)  
  }
ring.ID <- names(rings)

## transpose rings for easier aggregation
rings.t <- t(rings)

## recode ring IDS to average multiple cores from same tree
recode <- substr(as.character(colnames(rings)),1,5)
rings.recode <- data.frame(recode,rings.t)

## ring averages per tree
rings.agg <- aggregate(rings.recode,by=list(recode),FUN=mean,na.rm=TRUE)

## match rings and treering plot data
## ems treerings are also usable - have not looked at format recently, but is hopefully similar
lyford.plots <- read.csv("Data/Tree_Rings/Lyford_Data_13m/LyfordAllPlots.csv",skip=2)

## append site & tree number
rings.agg$Site <- substr(as.character(rings.agg$Group.1),1,3)
rings.agg$Tree.Number <- as.numeric(as.character(substr(as.character(rings.agg$Group.1),4,6)))

## merge plot and treering
rings.plot.merge <- merge(rings.agg,lyford.plots,by=c("Site","Tree.Number"))

## "Tag" from rings.plot.merge and "tree" from hf-c-trees-live should match
## data of live trees are in long, not wide format
## subset just lyford
lyf <- which(tree$site == "lyford")
lyford.tree <- tree[lyf,]

```

## 2. Estimate any uncertainties required as inputs, but which were not provided in the data

### 2a. Load Allometric posteriors
```{r}
b0 = -2.48
b1 = 2.48
B = c(b0,b1)
SIGMA = solve(diag(0.0001,2))
tau.allom = 100000
```


## 3. Format data for JAGS

```{r}

year = tree$year
years = min(year):max(year)
nyears = length(years)
year.index = year+1-years[1]

site = as.factor(as.character(tree$site))
sites = unique(site)
nsites = length(sites)
site.index = as.integer(site)

plot = as.factor(paste0(as.character(site),"+",as.character(tree$plot)))
plots = unique(plot)
nplots = length(plots)
plot.index = as.integer(plot)

s2p = table(site,plot)

area = tree$size
plot.area = tapply(area,plot,mean,na.rm=TRUE)
plot.yr.cnt = table(plot,year)
stem.count = apply(plot.yr.cnt,2,sum,na.rm=TRUE)
plot.yr.area = plot.yr.cnt
for(y in 1:ncol(plot.yr.area)){
  sel = which(plot.yr.area[,y]>0)
  plot.yr.area[sel,y] = plot.area[sel]
}
census.area = apply(plot.yr.area,2,sum,na.rm=TRUE)
census.year = as.numeric(names(census.area))-years[1]+1

## need to break out census.area & census.time by SITE
site.yr.cnt = site.yr.area = matrix(NA,nrow(s2p),ncol(plot.yr.area))
mch = match(colnames(s2p),rownames(plot.yr.area))
s2pv = apply(s2p,2,function(s){which(as.logical(s))})
for(y in 1:ncol(plot.yr.area)){
  site.yr.area[,y] = tapply(plot.yr.area[,y],s2pv,sum)
  site.yr.cnt[,y]  = tapply(plot.yr.cnt[,y],s2pv,sum)
}
#site.yr.area[is.na(site.yr.cnt)] <- NA
#site.yr.cnt[is.na(site.yr.cnt)] <- NA
site.yr.time = which(site.yr.cnt > 0,arr.ind = TRUE)

## empirical estimate of stem density
rho = site.yr.cnt/site.yr.area;rho[is.nan(rho)] <- NA
plot(rho[1,],ylim=range(rho,na.rm=TRUE),type='l')
for(s in 2:nsites){
  lines(rho[s,])
}

## empirical estimate of mean tree biomass
tree.mass <- exp(b0+b1*log(tree$dbh)+0.5/tau.allom)  ## empircal mean estimate needs var/2 correction
site.mass <- matrix(NA,nsites,nyears)
for(y in 1:nyears){
  sel.y = which(tree$year == years[y])
  if(length(sel)>0){
    site.mass[,y] = tapply(tree.mass[sel.y],site[sel.y],mean,na.rm=TRUE)
  }
}
mass.yr = which(!is.na(site.mass),arr.ind = TRUE)


#data = list(dbh=tree$dbh,n=nrow(tree),ny=nyears,yi=year.index,stem.count=stem.count,census.area=census.area,census.time=census.time,B=B,SIGMA=SIGMA,tau.allom=tau.allom)

data = list(dbh=tree$dbh,n=nrow(tree),ny=nyears,yi=year.index,stem.count=site.yr.cnt,census.area=site.yr.area,census.time=site.yr.time,B=B,SIGMA=SIGMA,tau.allom=tau.allom,ns=nsites,nobs=nrow(site.yr.time),census.year=census.year,si=site.index)

```


## 4. Specify / Estimate Priors


## 5. Specify / Estimate Initial Conditions for MCMC
```{r}
x = tree$dbh
mu.ic = mean(x,na.rm=TRUE)
prec.ic = 1/var(x,na.rm=TRUE)
alpha.ic = mu.ic - tapply(x,year.index,mean,na.rm=TRUE)
tau_year.ic = 1/var(alpha.ic,na.rm=TRUE)
alpha.ic[as.numeric(names(alpha.ic))] <- alpha.ic
alpha.ic[is.na(alpha.ic)] <- rnorm(sum(is.na(alpha.ic)),0,sqrt(1/tau_year.ic))

rhop.ic = apply(rho,1,mean,na.rm=TRUE)
tau_h.ic = 1/var(rhop.ic)
rho.ic = mean(rhop.ic)
tau_proc.rhop.ic = 1/mean(apply(rho,1,function(x){var(diff(x),na.rm=TRUE)}),na.rm=TRUE)

mup0.ic = apply(site.mass,1,mean,na.rm=TRUE)
tau_proc.mup.ic = 1/mean(apply(site.mass,1,function(x){var(diff(x),na.rm=TRUE)}),na.rm=TRUE)

init =  list(mu=mu.ic+alpha.ic,mup0 = mup0.ic,tau_proc.mup=tau_proc.mup.ic,tau_obs=prec.ic,tau_proc.mu=tau_year.ic,rhop0=rhop.ic,rho0=rho.ic,tau_proc.rhop=tau_proc.rhop.ic,tau_h.rho=tau_h.ic)
```

## 6. Basic Code

```{r}
BasicBudget = "
data{
  b ~ dmnorm(B,SIGMA)
  for(i in 1:n){
    mu.allom[i] <- b[1] + b[2]*log(dbh[i]) ## expected allometric biomass, log scale
    Emass[i] ~ dlnorm(mu.allom[i],tau.allom) 
  }
}
model{

  ## Stem Biomass #################
  mu0 ~ dlnorm(3,0.001)
  for(i in 1:ns){
    mup0[i] ~dlnorm(3,0.001) 
  }
  tau_proc.mu  ~ dgamma(0.1,0.1)
  tau_proc.mup ~ dgamma(0.1,0.1)
  tau_h.mu ~ dgamma(0.1,0.1)
  tau_obs  ~ dgamma(0.1,0.1)

  ## Process model: Landscape scale
  mu[1] <- mu0
  for(y in 2:ny){
    E[y] <- log(mu[y-1]) 
    mu[y] ~ dlnorm(E[y],tau_proc.mu)
  }

  ## Process model: Site scale
##  pmp <- tau_proc.mup+tau_h.mu
  pmp <- tau_proc.mup               ## simplified that drop's hier
  for(s in 1:ns){
##    Emp[s,1] <- log(mup0[s])*tau_proc.mup/pmp + log(mu[1])*tau_h.mu/pmp
#    Emp[s,1] <- log(mup0[s])
#    mup[s,1] ~ dlnorm(Emp[s,1],pmp)
    mup[s,1] <- mup0[s]
    for(y in 2:ny){
##      Emp[s,y] <- log(mup[s,y-1])*tau_proc.mup/pmp + log(mu[y])*tau_h.mu/pmp
      Emp[s,y] <- log(mup[s,y-1])
      mup[s,y] ~ dlnorm(Emp[s,y],pmp)
      }
  }

  ## observation model
  for(i in 1:n){
    theta[i] <- mup[si[i],yi[i]]
    Emass[i] ~ dnorm(theta[i],tau_obs)
  }

  ## Stem Density ################
  rho0 ~ dlnorm(-2,0.001)
  for(i in 1:ns){
    rhop0[i] ~ dlnorm(-2,0.001)
  }
  tau_proc.rho ~ dgamma(0.1,0.1)  # process landscale     ### ending up too tight
  tau_proc.rhop ~ dgamma(5,0.0005) # process within site              ### HACK
  tau_h.rho ~ dgamma(0.1,0.1)     # heterogeneity among sites

  ## State space latent rho
  rho[1] <- rho0
  for(y in 2:ny){
    Er[y] <- log(rho[y-1]) 
    rho[y] ~ dlnorm(Er[y],tau_proc.rho)
  }
## seems to be missing any rhop

## need some way to represent
## rhop[s,y] ~ dlnorm(log(rho[y]),tau_h.rho)
## rhop[s,y] ~ dlnorm(log(rhop[s,y]),tau_proc.rhop)

  ## state space latent rhop
  prp <- tau_proc.rhop+tau_h.rho
  for(s in 1:ns){
    Erp[s,1] <- log(rhop0[s])*tau_proc.rhop/prp + log(rho[1])*tau_h.rho/prp
    rhop[s,1] ~ dlnorm(Erp[s,1],prp)
    for(y in 2:ny){
      Erp[s,y] <- log(rhop[s,y-1])*tau_proc.rhop/prp + log(rho[y])*tau_h.rho/prp #E[rhop];  INELEGANT SOL'N
      rhop[s,y] ~ dlnorm(Erp[s,y],prp)
      }
  }

## should the effect of rho on rhop be a MA? A site random effect? An AR1? 
## Expect anomalies to persist  *******

  ## observation model
  for(i in 1:nobs){
    Ecount[i] <- rhop[census.time[i,1],census.year[census.time[i,2]]]*census.area[census.time[i,1],census.time[i,2]]
##    Ecount[i] <- rho[census.year[census.time[i,2]]]*census.area[census.time[i,1],census.time[i,2]]
    stem.count[census.time[i,1],census.time[i,2]] ~ dpois(Ecount[i])
  }

}
"

## compile JAGS model
j.model   <- jags.model (file = textConnection(BasicBudget),
                           data = data,
                           inits = init,
                           n.chains = 3)

```


```{r}
## burn-in
bmcmc   <- coda.samples(model = j.model,
                              variable.names = c("mu","mup","tau_obs","tau_proc.mu","tau_proc.mup","rho","tau_proc.rho","tau_h.rho","tau_proc.rhop","rhop"),
                              n.iter = 10000)
save(bmcmc,file="~/documents/BasicBudget.RData")
```


## 7. Post-Processing & Convergence Checking

```{r}
plot(bmcmc)  		## mcmc history and density plot
autocorr.plot(bmcmc)		## autocorrelation
cumuplot(bmcmc)		## quantile plot
gelman.plot(bmcmc)		## GRB statistic
summary(bmcmc)		## summary table
mu = as.data.frame(as.matrix(bmcmc))$mu
quantile(mu,c(0.025,0.5,0.975))
```

## 8. Visualization

```{r}
b.df = as.data.frame(as.matrix(bmcmc))
sel = 5000:10000
sel = c(sel,sel+10000,sel+20000)

cols = col2rgb(1:nsites,TRUE)
cols['alpha',]=cols['alpha',]/2
mycol <- function(i){rgb(cols[1,i],cols[2,i],cols[3,i],cols[4,i],maxColorValue=255)}

#### MEAN STEM BIOMASS ###################
ci = matrix(NA,nyears,3)
mup.ci = array(NA,c(nsites,nyears,3))
for(y in 1:nyears){
  ci[y,] <- quantile(b.df[sel,paste0("mu[",y,"]")],c(0.025,0.5,0.975))
  for(s in 1:nsites){
    mup.ci[s,y,] = quantile(b.df[sel,paste0("mup[",s,",",y,"]")],c(0.025,0.5,0.975))
  }
}
#yrng=range(ci); yrng[2] = min(yrng[2],200)
yrng = c(0,1000)
plot(years,ci[,2],type='n',ylim=yrng,ylab="Mean Tree Biomass (kg)",cex.lab=1.3)
#ciEnvelope(years,ci[,1],ci[,3],col="lightBlue")
#lines(years,ci[,2],lwd=3)
for(s in 1:nsites){
  ciEnvelope(years,mup.ci[s,,1],mup.ci[s,,3],col=mycol(s))
}
for(s in 1:nsites){
  lines(years,mup.ci[s,,2],lty=3,col=s,lwd=2)
  points(years,site.mass[s,],col=s,lty=2,lwd=3)
}

#### MEAN STEM DENSITY ###################
ci.r = matrix(NA,nyears,3)
rhop.ci = array(NA,c(nsites,nyears,3))
for(y in 1:nyears){
  ci.r[y,] <- quantile(b.df[sel,paste0("rho[",y,"]")],c(0.025,0.5,0.975))
  for(s in 1:nsites){
    rhop.ci[s,y,] = quantile(b.df[sel,paste0("rhop[",s,",",y,"]")],c(0.025,0.5,0.975))
  }
}
plot(years,ci.r[,2],type='n',ylim=c(0,0.3)#range(ci.r)
     ,ylab="Mean Stem Density",cex.lab=1.3)
#ciEnvelope(years,ci.r[,1],ci.r[,3],col="lightBlue")
#lines(years,ci.r[,2],lwd=3)
for(s in 1:nsites){
  ciEnvelope(years,rhop.ci[s,,1],rhop.ci[s,,3],col=mycol(s))
}
for(s in 1:nsites){
  lines(years,rhop.ci[s,,2],lty=3,col=s,lwd=2)
  points(census.year+years[1]-1,rho[s,],col=s,lty=2,lwd=3)
}
rho2 = matrix(NA,nsites,nyears)
for(s in 1:nsites){
  rho2[s,census.year] = rho[s,]
}

### ABOVEGROUND BIOMASS ################
ci.agb = matrix(NA,nyears,3)
AGBp.ci = array(NA,c(nsites,nyears,3))
for(y in 1:nyears){
  agb = b.df[sel,paste0("mu[",y,"]")]*b.df[sel,paste0("rho[",y,"]")]*10 # kg/m2 -> Mg/ha
  ci.agb[y,] <- quantile(agb,c(0.025,0.5,0.975))
  for(s in 1:nsites){
    AGBp.ci[s,y,] = quantile(b.df[sel,paste0("rhop[",s,",",y,"]")]*b.df[sel,paste0("mup[",s,",",y,"]")],c(0.025,0.5,0.975))*10
  }
}
agb.raw = rho2*site.mass*10
yrng=c(0,500)#range(AGBp.median)#range(ci); yrng[2] = min(yrng[2],150)
plot(years,ci.agb[,2],type='n',ylim=yrng,ylab="Mean AGB (Mg/ha)",cex.lab=1.3)
#ciEnvelope(years,ci.agb[,1],ci.agb[,3],col="lightBlue")
#lines(years,ci.agb[,2],lwd=3)
for(s in 1:nsites){
  ciEnvelope(years,AGBp.ci[s,,1],AGBp.ci[s,,3],col=mycol(s))
}
for(s in 1:nsites){
  lines(years,AGBp.ci[s,,2],lty=3,col=s,lwd=2)
  points(years,agb.raw[s,],col=s,lty=2,lwd=3)
}

## Plot the bivariate evolution of mean tree size and density
C=2/3*log(400)+log(0.09)
ylim=range(rhop.ci[,,2])
xlim=range(mup.ci[,,2])
mseq = seq(10,xlim[2],len=100)
n.thin = exp(C-2/3*log(mseq))
yr=years-years[1]+1
plot(mup.ci[1,,2],rhop.ci[1,,2],type='n',xlim=xlim,ylim=ylim,
    ylab="Stem Density (trees/m2)",xlab="Mean Tree Biomass (kg)",cex.lab=1.3)
for(i in 1:nsites){
  arrows(mup.ci[i,-nyears,2],rhop.ci[i,-nyears,2],mup.ci[i,-1,2],rhop.ci[i,-1,2],col=i,length=0.1,lwd=2)
#  points(rho2[i,],site.mass[i,],col=i)
}
lines(mseq,n.thin,lty=2,lwd=2)

lines(mseq,100/(mseq*10),lty=3)
lines(mseq,200/(mseq*10),lty=3)
lines(mseq,300/(mseq*10),lty=3)
lines(mseq,400/(mseq*10),lty=3)

```

## Composite AGB

```{r}
n0 = 2

## priors
mu0 = 250
tau0  = 0.000001
t1 = t2 = 0.01

## storage
ngibbs = nrow(b.df)
mu_h = tau_h = matrix(numeric(0),ngibbs,nyears)

## initial conditions
AGBsite = init$mup0*init$rho0*10
mu_h[1,] = mean(AGBsite)
tau_h[1,] = 1/var(AGBsite)

## MCMC
for(g in n0:ngibbs){
  
  for(y in 1:nyears){
    
    ## site AGB estimates
    AGB = b.df[g,paste0("mup[",1:nsites,",",y,"]")]*b.df[g,paste0("rhop[",1:nsites,",",y,"]")]*10
      
    ## update means
    w = nsites*tau_h[g-1,y]+tau0
    sd = sqrt(1/w)
    mu = (sum(AGB)*tau_h[g-1,y]+mu0*tau0)/w
    mu_h[g,y] = rnorm(1,mu,sd)
    
    ## update sd
    u1 <- t1 + nsites/2
    u2 <- t2 + 0.5*sum((AGB-mu_h[g,y])^2)
    tau_h[g,y] <- rgamma(1,u1,u2)    
  }
  
  if(g %% 100 == 0){print(g)}
}

save(bmcmc,mu_h,tau_h,file="~/Documents/BasicBudget.RData")
```

```{r}
### ABOVEGROUND BIOMASS ################
sel = 5000:10000
sel = c(sel,sel+10000,sel+20000)
ci.agb = matrix(NA,nyears,3)
for(y in 1:nyears){
  ci.agb[y,] <- quantile(mu_h[sel,y],c(0.025,0.5,0.975),na.rm=TRUE)
}
#agb.raw = rho2*site.mass*10
yrng=c(0,500)#range(AGBp.median)#range(ci); yrng[2] = min(yrng[2],150)
plot(years,ci.agb[,2],type='n',ylim=yrng,ylab="Mean AGB (Mg/ha)",cex.lab=1.3)
ciEnvelope(years,ci.agb[,1],ci.agb[,3],col="lightBlue")
lines(years,ci.agb[,2],lwd=6,col="darkBlue")
for(s in 1:nsites){
  lines(years,AGBp.ci[s,,2],lty=3,col=s,lwd=2)
  points(years,agb.raw[s,],col=s,lty=2,lwd=3)
}


```


